---
title: "DS 202 Final Project"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggthemes)

```

## Introduction/Background:

Our group wanted to take an in-depth look at American work-life balance. There are many facets to explore, such as how sex, job type, marital status, location, age, and number of dependents play a role in work-life balance, and how this has changed over time.

We collected our data from the IPUMS database, which has a collection of American Census data as well as American Community Surveys. The Census records span from 1790-2010, and the American Community Surveys span from 2000-present. The IPUMS database provides free access to social and economic data, and there is a massive collection of observations, with a wide array of variables.  



## Data Collection:
On IPUMS database, we selected a subset of variables and then selected a sample of 15,000 households per census/survey. We narrowed our scope to be from 1980-now, to get a better look at the most recent changes in work-life balance.

## Questions

### How has the family breadwinner role changed over time?

### How does state affect the relationship between work/life balance?

### How does gender play a role in work/life balance

### How many hours does the average American work per week? How as that changed over time?

### What types of jobs typically require longer commutes and longer work weeks?

### How does the American work life balance change in regards to income, marriage rates, and kids or dependents?  
```{r}
data2 <- data %>%
  select(UHRSWORK, NCHILD, INCTOT, FTOTINC,MARST, YEAR, TRANTIME) %>%
  rename(hours = UHRSWORK,
         dependents = NCHILD,
         incomesolo = INCTOT,
         incomefamily = FTOTINC,
         marital = MARST,
         year = YEAR,
         traintime = TRANTIME)
data2 <- na.omit(data2)
data2$married <- ifelse(data2$marital<=2, 1, 0) # 1 married, 0 not married
ggplot(data2, aes(y = hours, x = marital, fill=as_factor(marital)))+geom_boxplot()
ggplot(data2, aes(y = hours, x = dependents, fill=factor(dependents)))+geom_boxplot()
```
  
When we built the data set we initially tried to ad variables we thought would be important to determining work life balance, measured in hours worked per week. These included variables like marital status, family income, family size in regards to defendants, as well as other variables we thought may have been useful predictors. The two Plots I choose to explain this problem are above. These two plots are things that we thought would be very influential variables to determine work life balance but ended up having very little impact after all but ended up being less explanatory than we had predicted. They do a good job showing how little difference can be created as variables are changed. In the end we ended up finding that although the predictors we picked are valuable we discovered that they are not as valuable as we thought. There are probably many reliant predictors that we did not have access to that would have been more valuable but that is outside the scope of this analysis.

### What variables in our data set are significantly changed overtime when we predict income or family income?

### Are there any changes in the economic center cities and their proportion? Is this related to the changes in their life and work balance?
