---
title: "DS 202 Final Project"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(ipumsr)

ddi <- read_ipums_ddi("usa_00002.xml")
data <- read_ipums_micro(ddi)

```

## Introduction/Background:

Our group wanted to take an in-depth look at American work-life balance. There are many facets to explore, such as how sex, job type, marital status, location, age, and number of dependents play a role in work-life balance, and how this has changed over time.

We collected our data from the IPUMS database, which has a collection of American Census data as well as American Community Surveys. The Census records span from 1790-2010, and the American Community Surveys span from 2000-present. The IPUMS database provides free access to social and economic data, and there is a massive collection of observations, with a wide array of variables.  



## Data Collection:
On IPUMS database, we selected a subset of variables and then selected a sample of 15,000 households per census/survey. We narrowed our scope to be from 1980-now, to get a better look at the most recent changes in work-life balance.

## Questions
<> Start Example
<> TODO: Remove this block
    
    ```{r}
    data %>%
      filter(CLASSWKR > 0) %>%
      mutate(UHRSWORK = ifelse(YEAR == 2000, UHRSWORK / 2, UHRSWORK)) %>%
      ggplot(aes(x=YEAR, y=UHRSWORK)) + geom_bar(fill = "deepskyblue", stat="identity") +
      labs(x="Year", y= "Total Hours Worked per week", title="Hours of Work Per Week per Year") +
      geom_vline(xintercept=2009, color="red") +
      annotate("label", x=2008, y=7.5e5, label="Great Recession", hjust=0, fill="red", color="white")
    ```

<> End Example


### How has the family breadwinner role changed over time?

### How does state affect the relationship between work/life balance?

### How does gender play a role in work/life balance

### How many hours does the average American work per week? How as that changed over time?

### What types of jobs typically require longer commutes and longer work weeks?

### How does the American work life balance change in regards to income, marriage rates, and kids or dependents?  

### What variables in our data set are significantly changed overtime when we predict income or family income?

### Are there any changes in the economic center cities and their proportion? Is this related to the changes in their life and work balance?
