---
title: "Proposal"
author: "Kerstin Swanson, John Barmen, Braedon Giblin, Yufei Huang, Emma Smykalski"
date: "3/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Team members who actively participated:
John Barmen  
Kerstin Swanson  
Braedon Giblin  
Yufei Huang  
Emma Smykalski  


## Link to dataset



## Overview of dataset

```{r}

```





## Plan for exploratory analysis

Questions:  
-- How has the family breadwinner role changed over time?  
-- How does state effect the relationship between work/life balance? (Factoring in time commuting to and from work.) 
-- Does gender play a role in work/life balance?  
-- How is happiness effected by hours worked and commute time?  
-- How has the American work life balance changed in regards to income, marriage rates, and kids or dependents?  
-- What variables in our data set are significantly changed overtime when we predict income or family income?  
-- Are there any changes in the economic center cities and their proportion? Is this related to the changes in their life and work balance?  



