---
title: "Proposal"
author: "Kerstin Swanson, John Barmen, Braedon Giblin, Yufei Huang, Emma Smykalski"
date: "3/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Team members who actively participated:
John Barmen  
Kerstin Swanson  
Braedon Giblin  
Yufei Huang  
Emma Smykalski  


## Link to dataset
Since we curated our own variables, this is the link to our specific dataset:
https://github.com/JohnBarmen/DS202_FinalProject/blob/main/usa_00002.dat

Full data website:
https://usa.ipums.org/usa/index.shtml

## Overview of dataset

```{r}
library(ipumsr)
ddi <- read_ipums_ddi("usa_00002.xml")
data <- read_ipums_micro(ddi)
```


We collected data from IPUMS USA, which has a wide array of data from past censuses and American 
Community Surveys. The data ranges from 1790 to 2010. We selected a subset of variables to focus on our topic of Work/Life Balance in America.


We chose variables (between the dates of 1980-present) such as age, sex, income, state, number of children, marital status, hours spent per week at work, transportation time to work, job classifcation, and more. 



https://ipums.org/projects/ipums-usa

Sarah Flood, Miriam King, Renae Rodgers, Steven Ruggles and J. Robert Warren. Integrated Public Use Microdata Series, Current Population Survey: Version 8.0 [dataset]. Minneapolis, MN: IPUMS, 2020. https://doi.org/10.18128/D030.V8.0


## Plan for exploratory analysis

Questions:  
-- How has the family breadwinner role changed over time?  
-- How does state affect the relationship between work/life balance? (Factoring in time commuting to and from work.) 
-- Does gender play a role in work/life balance?  
-- How many hours does the average American work per week? How has that changed over time? 
-- What types of jobs typically require longer commutes and longer work weeks?
-- How has the American work life balance changed in regards to income, marriage rates, and kids or dependents?  
-- What variables in our data set are significantly changed overtime when we predict income or family income?  
-- Are there any changes in the economic center cities and their proportion? Is this related to the changes in their life and work balance?  



